---
title: "density"
output: html_document
---

```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling
library(tigris)
library(sp)

options(tigris_year = 2019)
```

```{r}
df_clarke <- tracts(state = 'GA', county = c('Clarke'))
plot(df_clarke)

df <- tracts(state = 'GA', county = c('Clarke', 'Oconee'))
```
The GEOID in the df dataframe matches our FIPS code for the population/sample data. Let's add the land area variable to the tractpopdata dataframe and then mutate a new variable to calculate population density. Then, we can compare population density to MP concentration. 

```{r}
pop <- readRDS(here("data", "processed_data", "tractpopdata.rds"))

area <- df %>% select(GEOID, ALAND)

combine <- left_join(pop, area, by = c("FIPS" = "GEOID"))

#calculate pop density
combine <- combine %>% mutate(density = population/ALAND)

# plot MP concentration vs pop density

combine %>% ggplot(aes(x = density, y = particles_l)) + geom_point() +
  labs(title = "Microplastic Concentration vs Population Density", x = "Population Density", y = "Particles/Liter") + scale_y_log10() + scale_x_log10()

```

# use linear regression to confirm relationship
```{r}
#linear regression model specification set up
lm_mod <- linear_reg() %>% set_engine("lm")
#estimating/training linear model
lm_fit <- lm_mod %>% fit(particles_l ~ density, data = combine)
#save table
lm_fittable <- tidy(lm_fit)
#view summary of fit
(lm_fit)
summary(lm_fit)
#save file
#saveRDS(lm_fit_dup_table, file = here("results", "lm_fit_dup_table.rds"))

model = lm(particles_l ~ density, data = combine)
summary(model)
```

#calculate 60th percentile
```{r}
quantile(sampledata$particles_l, .60)
```

#land cover distribution
```{r}
land_long <- athensland %>% pivot_longer(12:26)
  
land_long %>% ggplot(aes(x = name, y = value)) + geom_col() + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

