---
title: "exploration"
output: html_document
---

For this project, I began with microplastic count data for various sample sites, variables including site, watershed, latitude, longitude, and a_count and b_count. So far, I have obtained the following additional data: 

 - local land cover based on coordinates converted to census FIPS codes, using NGIS data
 - bacterial data from corresponding UOWN sample dates/sites
 - location population levels by zip code, converted from lat/long and joined with census data. 
 
In this document, I will explore relationships between some of these variables. 

Quantities of main interest: 
 - particles per liter of microplastics
 - season/sample date
 - population level
 - proximity to water reclamation facility effluent
 - bacteria levels at corresponding samples sites/dates
 
# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)


#path to data
data_location <- here::here("data","processed_data","popdata.rds")

#load data
mpdata <- readRDS(data_location)

```

# take a look at data
```{r}
summary(mpdata)

# view particles/L by season

mpdata %>% ggplot(aes(x = month, y = particles_l)) + geom_boxplot(aes()) + labs(title = "Microplastic Concentration")


```
 
 
