---
title: "censusprocessing"
output: html_document
---

At this point, I have written a few processing files loading the microplastics data, land cover/use data, and UOWN data for this project. I'd also like to look at small area population. I think I can achieve this by looking at population by zip code in Athens Clarke County, which is available from the US Census. My goal is to convert the latitude and longitude coordinates for each sample site into a zip code, and then examine whether microplastic levels vary based on local population size/population density. 

The data will come from the American Community Survey 5-year estimate 2019, and I will plan to use API. To convert sample coordinates to comparable zip codes, I can use the `tidygeocoder` package. 

Citation for zipcode population data (saving for later): 
United States Census Bureau. B01001 SEX BY AGE, 2019 American Community Survey 5-Year Estimates. U.S. Census Bureau, American Community Survey Office. Web. 10 December 2020. http://www.census.gov/.

Downloaded at: https://www.georgia-demographics.com/zip_codes_by_population

```{r}
#load  packages
library(readxl) #for loading Excel files
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(revgeo) #for reverse geocoding coordinates to zipcode


#load data. 
combodata <- readRDS(here("data", "processed_data", "combodata.rds"))
```



# renaming variable to differentiate use in processing
```{r}
popdata <- combodata
#save data as RDS
saveRDS(popdata, file = here("data", "processed_data", "popdata.rds"))

```

Update 10/25/21

I am realizing that we can get more localized population levels by using census tracts, which are smaller levels of census geography. Let's try to use FIPS codes (which include tract info) to obtain population data that might be more informative. I've removed the previous code that converted coordinates to zip code and got population based on zip code. This code is now located in `ARCHIVE.Rmd`

# Loading data with FIPS variable, converted from coordinates

FIPS codes were obtained through a census API run on the local computer, then saved locally and copied into the project directory under `raw_data` for this analysis. This ensures that the code will be reproducible and consistent even if the API changes in the future. 
```{r}
fips <- readRDS(here("data", "raw_data", "fips_copy.RDS"))
#saveRDS(fips, here("data", "processed_data", "fips.RDS"))
```

Here, after attemping to use `get_decennial()` we receive an error message suggesting to use NHGIS data since the most recent small census geography data is not available through `get_decennial()`.

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org and placed in "raw_data" for this R project. 
library(ipumsr)

#read in census tract level population data
tract <- read.csv(here("data","raw_data","nhgis0001_csv", "nhgis0001_ds244_20195_tract.csv"))

#clean up dataset to remove unneeded columns/variables
tract <- tract %>% filter(STATE == "Georgia") %>% select(1:3, 6:9, 12, 40, 43:46)

#Now, we have the FIPS codes for each sample site (down to the tract/block level), and we have population counts by FIPS code. 

#First, we need to remove the last four digits of the FIPS code in the fips dataset so that it will correspond with the #FIPS code in the tract dataset. We'll also need to remove the G from the beginning of the tract FIPS codes. 


#combine individual codes for state, county, tract to get long form FIPS code
a <- tract %>% mutate(FIPS = paste(STATEA, COUNTYA, sep = "")) %>%
  mutate(FIPS = paste(FIPS, TRACTA, sep = "0"))

#keep only the variables we need, FIPS code and population size (ALUBE001)
a <- a %>% select("FIPS", "ALUBE001")
#remove last four digits from fips codes associates with samples, so that they coordinate with population census tract FIPS codes
b <- fips %>% mutate(FIPS = (substr(FIPS,1,nchar(FIPS)-4)))
#combine datasets by FIPS code, so that population size is added to our main dataset based on corresponding FIPS code
blockpop <- left_join(b, a, by = "FIPS")

#We still have some missing values, but when I search for the FIPS code that are missing population levels, I am able to find them in the tract dataset. It looks like the FIPS code/GEOID are sometimes the same but sometimes differ by one digit (usually a zero inserted in the middle of the string somewhere). Let's adjust the GEOID variable to make it comparable to the FIPS codes we are still missing population data for, and then we'll join the datasets again to try to fill in missing values. 

#let's try GEOID minus first 7 digits
tract$GEOID <- gsub("^.{0,7}", "", tract$GEOID) 
#selecting only the variables we want to add in our final combination dataset
tractpop <- tract %>% select("GEOID", "ALUBE001")
#joining based on FIPS code to fill in missing population data
c <- left_join(blockpop, tractpop, by = c("FIPS" = "GEOID"))

#This last attempt was successful and we now have a column that gives us the census tract level population count, with no missing values. We no longer need the first column, or the zip code and zip code population data that we looked at initially. Let's clean up this dataset for further exploration

#Remove first ALUBE001 column and zip, population columns
tractpopdata <- c %>% select(-"ALUBE001.x")
tractpopdata <- rename(tractpopdata, "population" = "ALUBE001.y")


#save as RDS
saveRDS(tractpopdata, file = here("data", "processed_data", "tractpopdata.rds"))

```

