---
title: "wrfprocessing"
output: html_document
---

I also wanted to look at proximity of sample sites to water reclamation facilities (WRFs). Let's try that. 
# load WRF location
```{r}
library(here)
library(geodist) #to calculate distance from sample site to each water reclamation facility
library(dplyr) #for data processing
library(tidyverse)

#load MP data
mpdata <- readRDS(here("data", "processed_data", "popdata.rds"))
#path to data
wrfdata_location <- here::here("data","processed_data","wrflocation.rds")

#load data
wrflocation <- readRDS(wrfdata_location)
```

Eventually, I would like to be able to view this information on a map - where WRFs are labelled, and samples sites are color-coded by density of microplastics found. For now, let's see if we can calculate distance otherwise. 

```{r}
#calculate distance between sample coordinates and WRF location
wrfdistance <- geodist_vec(mpdata$long, mpdata$lat, wrflocation$long, wrflocation$lat)
```

I think the product we have here is a matrix of the distance of each sample site from each of the three WRFs, in the order that they appear in the WRF data (Cedar Creek, North Oconee, Middle Oconee). But, the labelling of the data is not super clear, so I need to verify this to be sure. 


```{r}
test <- mpdata %>% mutate(wrfdist = geodist_vec(mpdata$long, mpdata$lat, wrflocation$long, wrflocation$lat))
```

This produced another column like I wanted, but I don't know which WRF the distance was calculated for. Maybe I need to save WRF locations as individual lat and long values, and produce three new columns that way? However, I am not sure what the final product would look like if we do that, because a sample site could be closely associated with one WRF but much further from another one. Maybe I need to narrow this down to proximity to the nearest WRF, so that each sample site is only associated with the distance to ONE WRF, not three. 

#Working on gathering WRF distance
```{r}
#set individual WRF locations and save as values
cedarcreeklat = 33.87351 
cedarcreeklong = -83.33639
northoconeelat = 33.93402
northoconeelong = -83.36018
middleoconeelat = 33.90862
middleoconeelong = -83.39214

mpdata <- mpdata %>% mutate(ccwrf = geodist_vec(mpdata$long, mpdata$lat, cedarcreeklong, cedarcreeklat))
mpdata <- mpdata %>% mutate(nowrf = geodist_vec(mpdata$long, mpdata$lat, northoconeelong, northoconeelat))
mpdata <- mpdata %>% mutate(miwrf = geodist_vec(mpdata$long, mpdata$lat, middleoconeelong, middleoconeelat))

#Since I only care about the proximity to the closest WRF, my next task is to write code that creates a new variable returning only the lowest value out of ccwrf, nowrf, and miwrf for each sample site. 

#subset data into just distance from each WRF
wrf <- mpdata %>% select("ccwrf", "nowrf", "miwrf", "site", "date") 
#find minimum distance to any WRF out of all three
val <- do.call(pmin, c(wrf, na.rm = TRUE)) 
minwrf <- transform(wrf, min=val, minCol = names(wrf)[max.col(wrf == val, 'first')])

#minwrf contains the variable min, which tells us the distance from the sample site to the nearest WRF. This is all of the information that we need, so now we will re-join to the original DF
mpwrfdata <- left_join(mpdata, minwrf)

#remove variables we won't need anymore
mpwrfdata <- mpwrfdata %>% select(-"ccwrf", 
                                  -"nowrf",
                                  -"miwrf")
#rename columns
mpwrfdata <- mpwrfdata %>% rename("dist" = "min", 
                     "wrf" = "minCol")

#save as RDS
saveRDS(mpwrfdata, file = here("data", "processed_data", "mpwrfdata.rds"))
```

Update 10/25/21
I've updated the population data by accessing census tract level population rather than zip code level population. This should give us more localized population levels and might paint a better picture of the relationship between MP particles per liter and local population level. 

#Adding WRF data to updated population data
```{r}
#load data
tractpopdata <- readRDS(here("data", "processed_data", "tractpopdata.rds"))

#add wrf data to tractpopdata
wrfjoin <- mpwrfdata %>% select("site", "date", "wrf", "dist") #selecting just the variables we want to add
mpwrfcombo <- left_join(tractpopdata, wrfjoin, by = c("site", "date")) #joining the two datasets

glimpse(mpwrfcombo)
#dist variable is saved as character when it needs to be numeric
#convert dist from character to numeric
mpwrfcombo <- mpwrfcombo %>% mutate(dist = as.numeric(dist))

#save as RDS
saveRDS(mpwrfcombo, file = here("data", "processed_data", "mpwrfcombo.rds"))
```

Now, `mpwrfcombo` contains the sample info and the new WRF distance info. However, something to work on going forward is that a sample site may be near a WRf but not downstream of it. It might be more important to know the distance to the nearest upstream WRF. I'll have to go back and consider this as I continue my exploration/analysis. 

Other than not including landcover data, `mpwrfcombo` is the most comprehensive dataset, including almost all of the data that I have compiled so far.  

